---
title: "MULTIVARIATE ANALYSIS OF THE WINEQUALITY DATASET"
author: "VAN TASI"
date: "2024-03-26"
output: html_document
---


Assess the association (correlation) between the variables of the dataset “winequality.white.”
```{r}
# Load libraries
library(readr)
library(dplyr)
```

```{r}
# Read in the wine dataset 
wine_data <- read.csv("C:/Users/user/OneDrive - United States International University (USIU)/SPRING 2024/STA 2010/Assign/winequality-white.csv", sep=";")

# Print the first few entries
cov(wine_data)

# Find the dimensions of the data
```
check data types
```{r}
# Check the column names
names(wine_data)

# Check data type/structure of each variable
str(wine_data)

# Change the Type variable data type
wine_data$quality <- as.factor(wine_data$quality)

# Check data type/structure again
str(wine_data)
```
```{r}
#summarizing correlations within the dataset
corr_matrix <- cor(wine_data[, sapply(wine_data, is.numeric)]) # For numeric columns
ggcorrplot::ggcorrplot(corr_matrix, hc.order = TRUE)

```

a. Select only three variables that are correlated. NOTE: The correlation may not be
strong but will be the highest of what is available.
```{r}
# Compute the correlation matrix from the wine data
cor.wine_data <- cor(wine_data[1:11])
correlation_matrix <- as.matrix(cor.wine_data)

# Exclude correlations between a variable and itself (diagonal elements)
diag(correlation_matrix) <- 0

# Compute absolute correlations
abs_correlation <- abs(correlation_matrix)

# Find the indices of the highest absolute correlations
indices <- which(abs_correlation == max(abs_correlation), arr.ind = TRUE)

# Get the variable names corresponding to the highest absolute correlations
pair_names <- c(rownames(indices), colnames(indices))

# Print the pair names
print(paste(pair_names[1], "-", pair_names[2], ":", abs(correlation_matrix[indices]), sep = " "))

# Remove the highest absolute correlation and repeat the process for the next highest
abs_correlation[indices] <- 0

# Find the indices of the second highest absolute correlations
indices <- which(abs_correlation == max(abs_correlation), arr.ind = TRUE)

# Get the variable names corresponding to the second highest absolute correlations
pair_names <- c(rownames(indices), colnames(indices))

# Print the pair names
print(paste(pair_names[1], "-", pair_names[2], ":", abs(correlation_matrix[indices]), sep = " "))

# Remove the second highest absolute correlation and repeat the process for the third highest
abs_correlation[indices] <- 0

# Find the indices of the third highest absolute correlations
indices <- which(abs_correlation == max(abs_correlation), arr.ind = TRUE)

# Get the variable names corresponding to the third highest absolute correlations
pair_names <- c(rownames(indices), colnames(indices))

# Print the pair names
print(paste(pair_names[1], "-", pair_names[2], ":", abs(correlation_matrix[indices]), sep = " "))
```
We used absolute values to get the highest correlation values and not necessarily the strongest correlation. The three variables chosen are density, residual.sugar and alcohol. This can be confirmed by the plot below
```{r}
# Produce a matrix of plots for the first four variables 
library(ggplot2)
library(GGally)
wine_data.gg <- ggpairs(data = wine_data, columns = c(4,8,11))
wine_data.gg

# Color the points by wine type 
wine_data.gg <- ggpairs(data = wine_data, mapping = aes(color = quality), columns = c(4,8,11))

wine_data.gg
```

Subset the wine data to work with the three chosen variables.
```{r}
# Subset the wine data to include only the chosen variables
chosen_var <- wine_data[, c("residual.sugar", "density", "alcohol")]

# Print the first few entries
head(chosen_var)
```

b. Fit a correlation plot for the selected variables.
```{r}
# Compute the correlation of the chosen variables
cor.chosen <- cor(chosen_var)
cor.chosen
# Plot the correlations 
corrplot::corrplot(cor.chosen, method = "ellipse")
```

c. Compute the mean vector and the variance-covariance matrix of the selected
variables.
mean vector
```{r}
# Calculate the mean of 3 variables 
colMeans(chosen_var[, 1:3])

# Calculate the mean of the variables by wine quality
by(chosen_var, wine_data$quality, colMeans)
```

covariance
```{r}
# Calculate the variance-covariance matrix of the variables residual.sugar, density and alcohol
cov.chosen <- cov(chosen_var)

# Print the variance-covariance matrix
print("Variance-Covariance Matrix:")
print(cov.chosen)
```

d. Get a subset of 100 records with the three selected variables.
```{r}
chosen_subset <- chosen_var[1:100, ]
chosen_subset
```
i. Calculate the density heights for a multivariate normal. Plot a 3D scatterplot
of the density heights at each of the generated sample points.
```{r}
library(scatterplot3d) 
library(mvtnorm)

# Calculate the mean vector and covariance matrix for the selected variables
mu_sim <- colMeans(chosen_subset)
sigma_sim <- cov(chosen_subset)

multnorm_sample <- rmvnorm(n = 100, mean = mu_sim, sigma = sigma_sim)

multnorm_dens <- dmvnorm(chosen_subset, mean = mu_sim, sigma = sigma_sim)

# Create scatter plot of density heights 
scatterplot3d(cbind(multnorm_sample, multnorm_dens),    
                color="blue", pch="", type = "h",             
                              xlab = "x", ylab = "y", zlab = "density")
```
```{r}

# Load the mvtnorm package for multivariate normal calculations
library(mvtnorm)
library(ggplot2)
library(plotly)

# Calculate the mean vector and covariance matrix for the selected variables
mu_sim <- colMeans(chosen_subset)
sigma_sim <- cov(chosen_subset)

# Probability level
p <- 0.95

# Function to compute contour lines
compute_contour <- function(mu, sigma, p, n = 100) {
  # Generate grid of points
  x <- seq(mu[1] - 3*sqrt(sigma[1, 1]), mu[1] + 3*sqrt(sigma[1, 1]), length.out = n)
  y <- seq(mu[2] - 3*sqrt(sigma[2, 2]), mu[2] + 3*sqrt(sigma[2, 2]), length.out = n)
  z <- seq(mu[3] - 3*sqrt(sigma[3, 3]), mu[3] + 3*sqrt(sigma[3, 3]), length.out = n)
  grid <- expand.grid(x = x, y = y, z = z)
  
  # Compute the multivariate normal density for each point in the grid
  density <- dmvnorm(grid, mean = mu, sigma = sigma)
  
  # Reshape density vector into a three-dimensional array
  dim(density) <- c(length(x), length(y), length(z))
  
  # Find contour levels corresponding to the desired probability
  contour_levels <- quantile(density, p)
  
  # Compute the contour lines
  contours <- list()
  contours$x <- x
  contours$y <- y
  contours$z <- z
  contours$density <- density
  contours$contour_levels <- contour_levels
  
  # Return the contour lines
  return(contours)
}

# Compute contour lines
contours <- compute_contour(mu_sim, sigma_sim, p)

# Create an interactive 3D plot with contour lines
plot_ly(x = contours$x, y = contours$y, z = contours$z, 
        type = "contour", contours = list(z = list(show = TRUE, usecolormap = TRUE, project = TRUE)),
        colorscale = list(c(0, "rgb(255,0,0)"), list(0.5, "rgb(0,255,0)"), c(1, "rgb(0,0,255)")))

```





```{r}
library(MVN)
subset_mvntest <- mvn(chosen_subset, mvnTest="mardia",multivariatePlot = "qq")
subset_mvntest
```

e. Generate 200 samples from the dataset containing the three variables you had
earlier.
```{r}
chosen_subset_two <- chosen_var[201:400,]
chosen_subset_two
```

f. Assuming a multivariate t-distribution, check whether the samples follow a
multivariate normal distribution using the Mardia Test and plot the relevant qqplot
for the test.
```{r}
sub_two_mvntest <- mvn(chosen_subset_two, mvnTest="mardia",multivariatePlot = "qq")
sub_two_mvntest
```
```{r}
print(sub_two_mvntest)


```





